Dark Control:\\ Towards a Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-12-07 paper/paper.tex
reread part 2

% \begin{wrapfigure}{r}{.4\linewidth}
\begin{figure}[!htbp]
  \begin{minipage}[c]{0.63\textwidth}
    \includegraphics[width=.8\linewidth]{fig_dmn.png}
  \end{minipage}
  \hspace{-4em}
  \begin{minipage}[c]{0.45\textwidth}
    \caption{\textbf{Morphological coupling between reward system and
        default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11 \pm 0.02$) and right NAc volume ($R^2=0.14 \pm 0.02$) could be predicted from volume in the DMN regions. These out-of-sample generalization performances were obtained from support vector regression applied to normalized region volumes in the DMN in a 10-fold cross-validation procedure.
      Consistent for the left and right reward system, NAc volume in a given subject is
      positively coupled with the vmPFC and HC. The colors are indicative of the (\textcolor{red}{red} = positive,
      \textcolor{blue}{blue} = negative) and relative importance (the lighter the higher) of the regression coefficients.
  Code for reproduction and visualization:
  \url{www.github.com/banilo/darkcontrol_PCB2018}.
  }
  \label{fig:dmn}
\end{minipage}
% \end{wrapfigure}
\end{figure}

@@

\subsection{The hippocampus: memory, space, and experience replay}
The DMN midline has close functional links
with the HC in the medial temporal lobe \citep{vincet2006, shannon2013morning} \textemdash
a region long known to be involved in
memory operations and spatial navigation in animals and humans.
While the HC
is traditionally believed to allow recalling past experience,
there is now increasing evidence for an important role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering, gelbard2008internally, Javadi2017,
boyer2008evolutionary}.
Its recursive anatomical architecture
may be specifically designed to allow reconstructing
entire sequences of experience from memory fragments.
Indeed,
hippocampal damage is
not only associated with an impairment in re-experiencing the past (i.e., amnesia),
but also imagination dedicated to one's own future and
imagination of experiences more broadly \citep{hassabis2007patients}.

Mental scenes created by neurological patients with HC lesion exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal HC revealed
constantly active neuronal populations whose firing coincided with
specific locations in space during environmental navigation.
Indeed, when an animal is choosing between alternative
paths, the corresponding neuronal populations in the HC
spike one after another  \citep{johnson2007neural}.
Such neuronal patterns in the HC appear to directly indicate upcoming behavior,
such as in planning navigational trajectories
\citep{pfeiffer2013hippocampal} and
memory consolidation of choice relevance \citep{lavilleon2015}.
Congruently,
London taxi drivers, humans with high performance in spatial navigation,
were shown to exhibit increased gray-matter volume in the
HC \citep{maguire2000navigation}.

There is hence increasing evidence that
HC function extends beyond simple forms of
encoding and reconstruction of memory and space information.
Based on spike recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, which were elicited by experimental task design,
have been shown to be re-enacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space, o2010play}.
Moreover, neuronal spike sequences measured in hippocampal place cells of rats
featured re-occurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Similar spiking patterns in hippocampal neurons during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the regions of the DMN.
Moreover, in mice, invasively triggering spatial experience recall
in the HC during sleep
has been demonstrated to subsequently alter action choice during wakefulness
\citep{lavilleon2015}.
These HC-subserved mechanisms
conceivably contribute to advanced cognitive processes that require
re-experiencing or newly constructed mental scenarios,
such as in recalling autobiographical memory episodes
\citep{hassabis2007patients}.
Thus, the HC would orchestrate re-experience of environmental aspects for
consolidations based on re-enactment and for integration into
rich mental scene construction \citep{deuker2016event, bird2010establishing}.
As such, the HC may impact
ongoing perception of and action on the environment
\citep{maguire2016, lavilleon2015}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

\subsection{The right and left TPJ: prediction error signaling and world semantics}
The DMN emerges with its midline structures early in human development
\citep{doria2010}, while
the right and left TPJs may become fully integrated into this major brain
network only after birth.
The TPJs are known to exhibit hemispheric differences
based on microanatomical properties and gyrification patterns
\citep{seghier2013angular}.
Globally, neuroscientific investigations on hemispheric functional specialization
have highlighted the right versus left cerebral hemisphere as dominant for
attentional versus semantic functions
\citep{seghier2013angular, bzdok2013tpj, bzdok2016left,
stephan2007mechanisms}.

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
It is central for action initiation during goal-directed psychological tasks and for
sensorimotor behavior by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN region was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes,
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates functional control of brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, neurological patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with unexpected environmental stimuli and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The TPJ in the left-hemispheric DMN (LTPJ),
in turn, has a close relationship to Wernicke's area
involved in semantic processes, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from speech interpretations in linguistics
and neuropsychology,
the LTPJ mediates access to and integration of world knowledge,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
Consistent with this view,
LTPJ damage in humans also entails problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal production of
verbalized thought (``language of the mind'') was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and mental imagery strongly draw on
re-assembling world knowledge.
Isolated building blocks of world structure get rebuilt
in internally constructed mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast future events.
%
The LTPJ may hence facilitate the automated environmental predictions
by incorporating experience-derived building blocks of world regularities
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms is not possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of the $Q$-value function. This warrants a learning process
and definition of a loss function.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective, experience replay can be manifested as the re-occurrence of neuron spiking sequences that have also occurred during specific prior actions and environmental states. The HC is a strong candidate for contributing to such neural reinstantiation of behavioral episodes as neuroscience experiments have repeatedly indicated in rats, mice, cats, rabbits, songbirds, and monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.

>>>>

% \begin{wrapfigure}{r}{.4\linewidth}
\begin{figure}[!htbp]
  \begin{minipage}[c]{0.63\textwidth}
    \includegraphics[width=.8\linewidth]{fig_dmn.png}
  \end{minipage}
  \hspace{-4em}
  \begin{minipage}[c]{0.45\textwidth}
    \caption{\textbf{Morphological coupling between reward system and
        default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11 \pm 0.02$) and right NAc volume ($R^2=0.14 \pm 0.02$) could be predicted from volume in the DMN regions. These out-of-sample generalization performances were obtained from support vector regression applied to normalized region volumes in the DMN in a 10-fold cross-validation procedure.
      Consistent for the left and right reward system, NAc volume in a given subject is
      positively coupled with the vmPFC and HC. The colors are indicative of the (\textcolor{red}{red} = positive,
      \textcolor{blue}{blue} = negative) and relative importance (the lighter the higher) of the regression coefficients.
  The code for reproduction and visualization:
  \url{www.github.com/banilo/darkcontrol_PCB2018}.
  }
  \label{fig:dmn}
\end{minipage}
% \end{wrapfigure}
\end{figure}

@@

\subsection{The hippocampus: memory, space, and experience replay}
The DMN midline has close functional links
with the HC in the medial temporal lobe \citep{vincet2006, shannon2013morning} \textemdash
a region long known to be involved in
memory operations and spatial navigation in animals and humans.
While the HC
is traditionally believed to allow recalling past experience,
there is now increasing evidence for an important role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering, gelbard2008internally, Javadi2017,
boyer2008evolutionary}.
Its recursive anatomical architecture
may be specifically designed to allow reconstructing
entire sequences of experience from memory fragments.
Indeed,
hippocampal damage was
not only associated with an impairment in re-experiencing the past (i.e., amnesia),
but also forecasting of one's own future and
imagination of experiences more broadly \citep{hassabis2007patients}.

Mental scenes created by neurological patients with HC lesion exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal HC revealed
constantly active neuronal populations whose firing coincided with
specific locations in space during environmental navigation.
Indeed, when an animal is choosing between alternative
paths, the corresponding neuronal populations in the HC
spike one after another  \citep{johnson2007neural}.
Such neuronal patterns in the HC appear to directly indicate upcoming behavior,
such as in planning navigational trajectories
\citep{pfeiffer2013hippocampal} and
memory consolidation of choice relevance \citep{lavilleon2015}.
Congruently,
London taxi drivers, humans with high performance in forecasting spatial navigation,
were shown to exhibit increased gray-matter volume in the
HC \citep{maguire2000navigation}.

There is hence increasing evidence that
HC function extends beyond simple forms of
encoding and reconstruction of memory and space information.
Based on spike recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, which were elicited by experimental task design,
have been shown to be re-enacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space, o2010play}.
Moreover, neuronal spike sequences measured in hippocampal place cells of rats
featured re-occurrence directly after experimental trials
as well as directly before (prediction of) upcoming experimental trials \citep{diba2007forward}.
Similar spiking patterns in hippocampal neurons during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially including DMN regions.
Moreover, in mice, invasively triggering spatial experience recall
in the HC during sleep
has been demonstrated to subsequently alter action choice during wakefulness
\citep{lavilleon2015}.
These HC-subserved mechanisms
conceivably contribute to advanced cognitive processes that require
re-experiencing or newly constructed mental scenarios,
such as in recalling autobiographical memory episodes
\citep{hassabis2007patients}.
Thus, the HC would orchestrate re-experience of environmental aspects for
consolidations based on re-enactment and for integration into
rich mental scene construction \citep{deuker2016event, bird2010establishing}.
As such, the HC may impact
ongoing perception of and action on the environment
\citep{maguire2016, lavilleon2015}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

\subsection{The right and left TPJ: prediction error signaling and world semantics}
The DMN emerges with its midline structures early in human development
\citep{doria2010}, while
the right and left TPJs may become fully integrated into this major brain
network only after birth.
The TPJs are known to exhibit hemispheric differences
based on microanatomical properties and cortical gyrification patterns
\citep{seghier2013angular}.
Globally, neuroscientific investigations on hemispheric functional specialization
have highlighted the right cerebral hemisphere as dominant for attentional functions
and the left side for semantic functions
\citep{seghier2013angular, bzdok2013tpj, bzdok2016left,
stephan2007mechanisms}.

The TPJ in the right hemisphere (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
This DMN region is probably central for
action initiation during goal-directed psychological tasks and for
sensorimotor behavior by integrating multi-sensory attention
\citep{corbetta2002control}.
Its involvement was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and predicted outcomes,
without necessarily relating these neural processes to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates functional control of brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, neurological patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events, the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with unexpected environmental stimuli and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The left TPJ of the DMN (LTPJ),
in turn, has a close relationship to Wernicke's area
involved in semantic processes, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from speech interpretations in linguistics
and neuropsychology,
the LTPJ appears to mediate access to and binding of world knowledge,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
Consistent with this view,
LTPJ damage in humans also entailed problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal production of
verbalized thought (``language of the mind'') was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and mental imagery to forecast future events strongly draw on
re-assembling world knowledge.
Isolated building blocks of world structure get rebuilt
in internally constructed mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast event outcomes.
%
Neural processes in the LTPJ may hence
contribute to the automated predictions of the environment
by incorporating experience-derived building blocks of world regularities
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms is not possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of the $Q$-value function. This warrants a learning process
and definition of a loss function.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective, experience replay can be manifested as the re-occurrence of neuron spiking sequences that have also been measured during specific prior actions and environmental states. The HC is a strong candidate for contributing to such neural reinstantiation of behavioral episodes as neuroscience experiments have repeatedly indicated in rats, mice, cats, rabbits, songbirds, and monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.
